Experiment 0 pilot

### 2018-07-18 experiments/e0/writeup/e0.Rmd
revisited in July 2018

---
title: "Experiment 0 pilot"
author: "Matt Green"
output:
  html_document:
    toc: yes
    <!--toc_float: true-->
    mathjax: default
    template: default
    code_folding: hide
    theme: default
---

@@

Knitr doc options.

```{r docOpts}
library(knitr)
knit_hooks$set(crop=hook_pdfcrop)
```

@@

```{r loadLibraries}
suppressPackageStartupMessages(library(MASS))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(library(plyr))
suppressPackageStartupMessages(library(reshape))
suppressPackageStartupMessages(library(lattice))
suppressPackageStartupMessages(library(LMERConvenienceFunctions))
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(lmerTest))
```

@@

```{r sourceCIFunctions}
source("CIfunctions.R")
```

# Preprocessing

@@

```{r getData}
if( file.exists('data.txt') ) 
	{
	d <- read.table("data.txt", header=FALSE) 
} else {
	system("cat ../experimentCode/output/*.dat > data.txt", wait=TRUE)
}
mynames <- list("gap", "target", "side", "vagueness", "pair", "Version", "Left",
								"Right", "Text", "Condition", "correctAnswer", "subject",
								"actualAnswer", "isCorrect", "rt")
names(d) <- mynames
d$obs <- 1:nrow(d)
d$trial <- 0
for (s in unique(d$subject)){
  d[d$subject==s, "trial"] <- 1:nrow(subset(d,subject==s))
}
# For some reason there is one rt of -6. Set it to NA
d[d$rt==-6,"rt"] <- NA
# make a df
d <- data.frame(
  obs = d$obs, 
  trial = d$trial, 
  subject = d$subject, 
  item = factor(d$pair, label = c("2:4", "3:5", "6:8", "7:9", "2:6", "7:3", "8:4", "5:9")),
  gap = factor(d$gap, levels = c("BigGap", "SmallGap"), labels = c("Big Gap", "Small Gap")), 
  quan = factor(d$target, levels = c("BigTarget", "SmallTarget"), labels = c("Big Target", "Small Target")), 
  vag = factor(d$vagueness, levels = c("Precise", "Vague"), labels = c("Crisp", "Vague")), 
  side = d$side, 
  err = 1 - d$isCorrect, 
  rt = d$rt)
d$subit <- "Not subitizable"
d$subit[d$item == "2:4"] <- "Subitizable"
d$subit[d$item == "3:5"] <- "Subitizable"
d$subit[d$item == "2:6"] <- "Subitizable"
d$subit[d$item == "7:3"] <- "Subitizable"
d$subit = as.factor(d$subit)
d <- subset(d, select = c("obs", "trial", "subject", "item", "gap", "quan", 
                          "vag", "subit", "side", "err", "rt"))
```

Size of the data is \Sexpr{nrow(d)} trials. 1 was removed for being minus 6 rt.
Remove erroneous response trials. There are \Sexpr{nrow(subset(d,err==1))} error trials.

```{r removeErrorTrials}
# by setting the RT to NA
d[d$err==1,"rt"] <- NA
```

Do some factors and contrasts.  Set subject and item as factors. Sum code vagueness and subitizability.

```{r postProcessData}
d$subject <- as.factor(d$subject)
d$item <- as.factor(d$item)
d$vag <- relevel(d$vag, ref="Vague")
contrasts(d$vag) <- contr.sum(2)/2
d$subit <- relevel(d$subit, ref="Subitizable")
contrasts(d$subit) <- contr.sum(2)/2
contrasts(d$quan) <- contr.sum(2)/2
contrasts(d$gap) <- contr.sum(2)/2
```

Check the contrasts.

```{r checkContrasts}
contrasts(d$vag)
contrasts(d$subit)
contrasts(d$quan)
contrasts(d$gap)
```

Add transformations of rt

```{r addTransform}
d$logRT <- log(d$rt)
d$recipRT <- -10000/d$rt
```

For the untrimmed data, show effects of transforming rt.

```{r plotDistributions, fig.width=4, fig.height=2}
xs = melt(d,  
					measure.vars = c("rt","logRT","recipRT"))
densityplot(
	~value|variable,
	data=xs,
	xlab=NULL,
	yaxt=F,
  scales=list(x=list(relation="free", rot=45), 
   						y=list(relation="free", at=NULL)),
	cex=.2,
	plot.points=FALSE,
	layout=c(3,1)
	)
```

Trim the data. This function doesn't work well if there are NA values in the response, so explicitly remove them first.

```{r trimData}
dummy <- subset(d, !is.na(rt))
data.trim <- perSubjectTrim.fnc(dummy, response="rt",subject="subject", trim=2.5)$data
```

Check the distribution of the trimmed data to see if a transformation is suggested.

```{r checkTrim, fig.width=4, fig.height=2}
xs = melt(data.trim,  
					measure.vars = c("rt","logRT","recipRT"))
densityplot(
	~value|variable,
	data=xs,
	xlab=NULL,
	yaxt=F,
  scales=list(x=list(relation="free", rot=45), 
   						y=list(relation="free", at=NULL)),
	cex=.2,
	plot.points=FALSE,
	layout=c(3,1)
	)
```

@@

```{r selectTrim}
dat <- data.trim
```

@@

```{r, lmers}
lmer2 <- lmer(data=dat, logRT~subit*vag + (subit*vag|subject) + (vag|item))
# lmer3 is closer to the original anova than lmer2
lmer3 <- lmer(data=dat, logRT~subit*vag+quan+gap + (subit*vag|subject) + (vag|item))
```

```{r, summaryLMER, results='asis'}
print(xtable(summary(lmer3)$coefficients, digits=3), type="html")
```

## subitizable

```{r, subitRT, results='asis'}
print(xtable(summary(lmer(data=dat, subset=subit=="Subitizable",logRT~vag+quan+gap +(vag|subject)+(vag|item)))$coefficients, digits=3), type='html')
```

## not subitizable

```{r, notsubitrT, results='asis'}
print(xtable(summary(lmer(data=dat, subset=subit=="Not subitizable", logRT~vag+quan+gap +(vag|subject)+(vag|item)))$coefficients, digits=3), type='html')
```

# Error rate analysis

```{r errorrates, results='asis'}
print(xtable(summary(
	glmer(data=d, err~vag*subit+quan+gap + (vag|subject) + (1|item), 
				family=binomial, 
				control = glmerControl(optimizer = "bobyqa")
				)
	)$coefficients,
	digits=3), 
	type='html')
```

# Plots of significant effects

```{r plot_sig, fig.width=7.5, fig.height=3}
a <- summarySEwithin(data=dat, measurevar="logRT", withinvars=c("vag","subit","item"), idvar="subject")
as <- subset(a,subit=="Subitizable")
as$item <- as.character(as$item)
as$item[as$item=="7:3"] <- "3:7"
as$item <- factor(as$item)
pd <- position_dodge(.05)
plot1 = ggplot(data=as, aes(y=logRT, x=item, ymin=logRT-ci, ymax=ci+logRT, fill=vag, group=vag)) + 
  geom_line(position=pd) +
  geom_errorbar(width=.2, position=pd)+
  geom_point(pch=21, size=4, position=pd) +
  scale_fill_grey(name=element_blank(),start=0,end=1)+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.key=element_blank()) +
  ggtitle("Response time")

a <- summarySEwithin(data=d, measurevar="err", withinvars=c("vag","subit","item"), idvar="subject")
as <- subset(a,subit=="Subitizable")
as$item <- as.character(as$item)
as$item[as$item=="7:3"] <- "3:7"
as$item <- factor(as$item)
pd <- position_dodge(.05)
plot2 = ggplot(data=as, aes(y=err, x=item, ymin=err-ci, ymax=ci+err, group=vag, fill=vag))+
  geom_line(position=pd) +
  geom_errorbar(width=.2, position=pd)+
  geom_point(pch=21, size=4, position=pd) +
  scale_fill_grey(name=element_blank(),start=0,end=1)+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.key=element_blank()) +
  ggtitle("Error rate")

grid.arrange(plot1, plot2, nrow=1, top="Subitizable only")
```

```{r barcharts, fig.width=9, fig.height=3}
bwtheme <- standard.theme("pdf", color=FALSE)
b1=barchart(data=ddply(dat,.(vag,subit),summarise, logRT=mean(logRT)), 
						logRT~subit, 
						group=vag, auto.key=list(space='right', cex=.7), par.settings=bwtheme)
b2=barchart(data=ddply(d,.(vag,subit),summarise, Error_Rate=mean(err)), 
						Error_Rate~subit, 
						group=vag, auto.key=list(space='right', cex=.7), par.settings=bwtheme)
grid.arrange(b1,b2,ncol=2)
```

Subitizability x Vagueness x Quantity

```{r plotMeansa, fig.width=7, fig.height=3}
a <- summarySEwithin(data=dat, measurevar="logRT", withinvars=c("vag","quan","subit"), idvar="subject")
pd <- position_dodge(.1)
	ggplot(
		a, 
		aes(y=logRT, x=vag, group=quan, shape=quan, col=quan, ymin=logRT-ci, ymax=ci+logRT)
	) + 
	geom_point(size=4, position=pd) + 
	geom_line(position=pd) + 
	facet_grid(~subit) + 
	geom_errorbar(width=.1, position=pd) + 
	theme_bw() + 
	theme(panel.grid.major=element_blank()) 
```

```{r qqplotLMER, fig.width=3, fig.height=3.5}
qqPlot(residuals(lmer2),cex=.5)
```

>>>>

---
title: "Experiment 0 pilot"
author: "Matt Green"
output:
  html_document:
    toc: yes
    toc_float: true
    mathjax: default
    template: default
    code_folding: show
    theme: default

---

@@

The pilot experiment used subitizable numbers as well as bigger numbers.
The instructions had the form: _Choose the square with {{few, many},{two, three, four, five, six, seven, eight, nine}} dots._
The variables were: subitizability; vagueness; target_size; gap_size; side (i.e., which side was the bigger number on?)

# Libraries and functions

```{r load knitr}
library(knitr)
```

@@

```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(dotwhisker)
library(grid)
library(gridExtra)
library(lme4)
# R CMD INSTALL broom.mixed # https://github.com/bbolker/broom.mixed
```

@@

```{r sourceCIFunctions}
source("CIfunctions.R") # for summarySEwithin etc
```

# Data

@@

```{r "make the data.txt file unless it's already there; read the data from file, add some vars and code the added vars"}

# make the data file from the individual subjects' data unless it's already there in the directory
if (!file.exists("data.txt")) {
  system("echo 'gap\ttarget\tside\tvagueness\tpair\tVersion\tLeft\tRight\tText\tCondition\tcorrectAnswer\tsubject\tactualAnswer\tisCorrect\trt' > data.txt")
  system("cat ../experimentCode/output/*.dat >> data.txt", wait=TRUE)
}

# read in the data from file
din <- read_delim("data.txt", delim="\t")

# make an R object copy of that original data file
original_data <- din

# add observation number for the whole data set
din$obs <- 1:nrow(din)

# add trial index for each subject
din$trial <- 0
for (s in unique(din$subject)){
  din[din$subject==s, "trial"] <- 1:nrow(subset(din,subject==s))
}
```

```{r"make a tibble with factors explicit"}
# make a df to carry forward
d <- tibble(
  obs = din$obs, 
  trial = din$trial, 
  subject = as.factor(din$subject), 
  item = factor(din$pair, label = c("2:4", "3:5", "6:8", "7:9", "2:6", "7:3", "8:4", "5:9")),
  gap = factor(din$gap, levels = c("BigGap", "SmallGap"), labels = c("Big Gap", "Small Gap")), 
  quan = factor(din$target, levels = c("BigTarget", "SmallTarget"), labels = c("Big Target", "Small Target")), 
  vag = factor(din$vagueness, levels = c("Precise", "Vague"), labels = c("Crisp", "Vague")), 
  subit = factor(ifelse(din$pair %in% c(1,2,5,6), "Subitizable", "Not subitizable")),
  side = factor(din$side), 
  text = factor(din$Text),
  err = 1 - din$isCorrect, 
  rt_raw = ifelse(din$rt >=0, din$rt, NA),
  rt_ms = ifelse(1-din$isCorrect==1, NA, ifelse(din$rt >=0, din$rt, NA)), # restricting rt to positive values (rt>=0) converted a score of -6 ms to NA
  rt_log = log( ifelse(1-din$isCorrect==1, NA, ifelse(din$rt >=0, din$rt, NA)) ),
  "rt_reciprocal[-1/rt]" = -1/ifelse(1-din$isCorrect==1, NA, ifelse(din$rt >=0, din$rt, NA)), # We multiplied reciprocal scores [1/RT] by -1 [-1/RT] to maintain the direction of effects compatible for the three variants, effectively converting speed [number of responses per ms] into 'rate of slowing' - A linear mixed model analysis of masked repetition priming, Kliegl, Masson & Richter, VISUAL COGNITION,2010,18(5),655--681 (p.~662)
  rt_reciprocal = -1000/ifelse(1-din$isCorrect==1, NA, ifelse(din$rt >=0, din$rt, NA))
)
d$c.vag = ifelse(d$vag=="Crisp", -.5, .5)
```

Size of the data is `r nrow(d)` trials. 1 was removed for being impossible ($-6~ms$). There are `r nrow(subset(d,err==1))` error trials. The variable rt_ms has NA in the case that an error was made. The variable rt_raw includes rt for error trials.

## Eyeball the RTs

### Uncapped, max of 20,000 ms

```{r fig.show='hide', warning=FALSE, echo=FALSE}
# all the (correct) RTs
summary(d$rt_ms)

data_for_plot <- gather(d, key=rt_metric, value=rt_value, rt_ms, rt_log, "rt_reciprocal[-1/rt]") %>%
  arrange(rt_value)

ga1untrimmed = ggplot(data_for_plot, aes(x=rep(1:1600,times=3), y=rt_value)) + 
  geom_point(na.rm=TRUE) + 
  facet_wrap(~factor(rt_metric, levels=c("rt_ms", "rt_log", "rt_reciprocal[-1/rt]")), scales="free_y") +
  xlab("observations, ordered by rt_value") + theme(aspect.ratio = 1, axis.text.x = element_text(angle = 45, hjust = 1))

ga2untrimmed = ggplot(data_for_plot, aes(rt_value)) +
  geom_density(na.rm=TRUE) +
  facet_wrap(~factor(rt_metric, levels=c("rt_ms", "rt_log", "rt_reciprocal[-1/rt]")), scales="free") + theme(aspect.ratio = 1, axis.text.x = element_text(angle = 45, hjust = 1))

myg1 <- grid.arrange(ga1untrimmed, ga2untrimmed, nrow=2, top = textGrob("Untrimmed",gp=gpar(fontsize=20,font=3)))
```

### Capped at 10,000 ms

```{r fig.show='hide', warning=FALSE, echo=FALSE}
# now with a simple cap imposed at 10000 ms, half-way through the maximum allotted time of 20000 (2 minute timeout)
summary(d$rt_ms[d$rt_ms<10000])

data_for_plot <- gather(d %>% filter(rt_ms<10000), key=rt_metric, value=rt_value, rt_ms, rt_log, "rt_reciprocal[-1/rt]") %>%
  arrange(rt_value)

ga1trimmed = ggplot(data_for_plot, aes(x=rep(1:1556,times=3), y=rt_value)) + 
  geom_point(na.rm=TRUE) + 
  facet_wrap(~factor(rt_metric, levels=c("rt_ms", "rt_log", "rt_reciprocal[-1/rt]")), scales="free_y") +
  xlab("observations, ordered by rt_value")+ theme(aspect.ratio = 1, axis.text.x = element_text(angle = 45, hjust = 1))

ga2trimmed = ggplot(data_for_plot, aes(rt_value)) +
  geom_density(na.rm=TRUE) +
  facet_wrap(~factor(rt_metric, levels=c("rt_ms", "rt_log", "rt_reciprocal[-1/rt]")), scales="free") + theme(aspect.ratio = 1, axis.text.x = element_text(angle = 45, hjust = 1))

myg2 <- grid.arrange(ga1trimmed, ga2trimmed, nrow=2, top = textGrob("Trimmed at 10,1000 ms",gp=gpar(fontsize=20,font=3)))
```

```{r, fig.width=10,fig.height=10}
grid.arrange(myg1, myg2, nrow=2)
```

@@

```{r selectTrim}
dat <- d %>% filter(rt_ms<10000)
```

@@

```{r, lmers}
lmer2 <- lmer(data=dat, rt_reciprocal~subit*vag + (vag|subject) + (vag|item))
# lmer3 is closer to the original anova than lmer2
lmer3 <- tidy(lme4::lmer(data=dat, rt_reciprocal~vag*subit+quan+gap + (vag+subit|subject) + (vag|item)))
dwplot(lmer3 %>% filter(group=="fixed"), vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2)) + xlab("Coefficient")
```

<!-- ```{r} -->
<!-- lmer3b <- lme4::lmer(data=dat, logRT~subit*vag+quan+gap + (subit*vag|subject) + (vag|item))  -->
<!-- lmer3b_tidy <- tidy(lmer3b) -->
<!-- lmer3b_tidy$model="foo" -->
<!-- lmer3b_tidy %>% filter(group=="fixed") %>% dwplot() -->
<!-- ``` -->

<!-- ```{r, summaryLMER, results='asis'} -->
<!-- print(xtable(summary(lmer3)$coefficients, digits=3), type="html") -->
<!-- ``` -->

<!-- ## subitizable -->

<!-- ```{r, subitRT, results='asis'} -->
<!-- print(xtable(summary(lmer(data=dat, subset=subit=="Subitizable",logRT~vag+quan+gap +(vag|subject)+(vag|item)))$coefficients, digits=3), type='html') -->
<!-- ``` -->

<!-- ## not subitizable -->

<!-- ```{r, notsubitrT, results='asis'} -->
<!-- print(xtable(summary(lmer(data=dat, subset=subit=="Not subitizable", logRT~vag+quan+gap +(vag|subject)+(vag|item)))$coefficients, digits=3), type='html') -->
<!-- ``` -->

<!-- # Error rate analysis -->

<!-- ```{r errorrates, results='asis'} -->
<!-- print(xtable(summary( -->
<!-- 	glmer(data=d, err~vag*subit+quan+gap + (vag|subject) + (1|item),  -->
<!-- 				family=binomial,  -->
<!-- 				control = glmerControl(optimizer = "bobyqa") -->
<!-- 				) -->
<!-- 	)$coefficients, -->
<!-- 	digits=3),  -->
<!-- 	type='html') -->
<!-- ``` -->

<!-- # Plots of significant effects -->

<!-- ```{r plot_sig, fig.width=7.5, fig.height=3} -->
<!-- a <- summarySEwithin(data=dat, measurevar="logRT", withinvars=c("vag","subit","item"), idvar="subject") -->
<!-- as <- subset(a,subit=="Subitizable") -->
<!-- as$item <- as.character(as$item) -->
<!-- as$item[as$item=="7:3"] <- "3:7" -->
<!-- as$item <- factor(as$item) -->
<!-- pd <- position_dodge(.05) -->
<!-- plot1 = ggplot(data=as, aes(y=logRT, x=item, ymin=logRT-ci, ymax=ci+logRT, fill=vag, group=vag)) +  -->
<!--   geom_line(position=pd) + -->
<!--   geom_errorbar(width=.2, position=pd)+ -->
<!--   geom_point(pch=21, size=4, position=pd) + -->
<!--   scale_fill_grey(name=element_blank(),start=0,end=1)+ -->
<!--   theme_bw()+ -->
<!--   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.key=element_blank()) + -->
<!--   ggtitle("Response time") -->

<!-- a <- summarySEwithin(data=d, measurevar="err", withinvars=c("vag","subit","item"), idvar="subject") -->
<!-- as <- subset(a,subit=="Subitizable") -->
<!-- as$item <- as.character(as$item) -->
<!-- as$item[as$item=="7:3"] <- "3:7" -->
<!-- as$item <- factor(as$item) -->
<!-- pd <- position_dodge(.05) -->
<!-- plot2 = ggplot(data=as, aes(y=err, x=item, ymin=err-ci, ymax=ci+err, group=vag, fill=vag))+ -->
<!--   geom_line(position=pd) + -->
<!--   geom_errorbar(width=.2, position=pd)+ -->
<!--   geom_point(pch=21, size=4, position=pd) + -->
<!--   scale_fill_grey(name=element_blank(),start=0,end=1)+ -->
<!--   theme_bw()+ -->
<!--   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.key=element_blank()) + -->
<!--   ggtitle("Error rate") -->

<!-- grid.arrange(plot1, plot2, nrow=1, top="Subitizable only") -->
<!-- ``` -->

<!-- ```{r barcharts, fig.width=9, fig.height=3} -->
<!-- bwtheme <- standard.theme("pdf", color=FALSE) -->
<!-- b1=barchart(data=ddply(dat,.(vag,subit),summarise, logRT=mean(logRT)),  -->
<!-- 						logRT~subit,  -->
<!-- 						group=vag, auto.key=list(space='right', cex=.7), par.settings=bwtheme) -->
<!-- b2=barchart(data=ddply(d,.(vag,subit),summarise, Error_Rate=mean(err)),  -->
<!-- 						Error_Rate~subit,  -->
<!-- 						group=vag, auto.key=list(space='right', cex=.7), par.settings=bwtheme) -->
<!-- grid.arrange(b1,b2,ncol=2) -->
<!-- ``` -->

<!-- Subitizability x Vagueness x Quantity -->

<!-- ```{r plotMeansa, fig.width=7, fig.height=3} -->
<!-- a <- summarySEwithin(data=dat, measurevar="logRT", withinvars=c("vag","quan","subit"), idvar="subject") -->
<!-- pd <- position_dodge(.1) -->
<!-- 	ggplot( -->
<!-- 		a,  -->
<!-- 		aes(y=logRT, x=vag, group=quan, shape=quan, col=quan, ymin=logRT-ci, ymax=ci+logRT) -->
<!-- 	) +  -->
<!-- 	geom_point(size=4, position=pd) +  -->
<!-- 	geom_line(position=pd) +  -->
<!-- 	facet_grid(~subit) +  -->
<!-- 	geom_errorbar(width=.1, position=pd) +  -->
<!-- 	theme_bw() +  -->
<!-- 	theme(panel.grid.major=element_blank())  -->
<!-- ``` -->

<!-- ```{r qqplotLMER, fig.width=3, fig.height=3.5} -->
<!-- qqPlot(residuals(lmer2),cex=.5) -->
<!-- ``` -->
